Unshuffling Permutations

### 2015-09-16 article.tex
Introduction hardness

\begin{proposition}
  \label{proposition:matching}
  Let $\pi \in S_{2n}$.
  Then,
  $\pi$ is the union of two order-isomorphic patterns
  if and only if
  there exists a perfect containment-free mathing
  $\mathcal{M}$ of $G_\pi$ such that
  $\LEFT(\mathcal{M}) \simeq \RIGHT(\mathcal{M})$.
\end{proposition}

The following easy lemma will prove extremely useful for simplifying the proof of
Proposition~\ref{proposition:hardness}.

\begin{lemma}
  \label{lemma:at most one edge monotone}
  Let $\pi = \pi_1 \, \sigma_1 \, \pi_2 \, \sigma_2 \, \pi_3$
  be a permutation with $n_1 = |\sigma_1| \geq 2$ and $n_2 = |\sigma_2| \geq 2$,
  and $\mathcal{M}$ be a containment-free perfect matching of $G_\pi$.
  The following assertions hold.
  \begin{enumerate}
    \item
    If $\sigma_1$ is increasing, $\sigma_2$ is decreasing and
    $\sigma_1$ is above $\sigma_2$
    (see Fig.~\ref{subfig:increasing before and aboce decreasing}),
    then $\mathcal{M}$ sontains at most one $(\sigma_1, \sigma_2)$-edge.
    \item
    If $\sigma_1$ is decreasing, $\sigma_2$ is increasing and
    $\sigma_1$ is below $\sigma_2$
    (see Fig.~\ref{subfig:decresing before and below increasing}),
    then $\mathcal{M}$ sontains at most one $(\sigma_1, \sigma_2)$-edge.
  \end{enumerate}
\end{lemma}

@@

\begin{proof}
  For the sake of clarity, we will that a bunch of positions $P$
  is \emph{above} (resp. \emph{below}) above another bunch of positions $P'$
  in permutation $\mu$
  if $\mu(x) > \mu(y)$ (resp. $\mu(x) < \mu(y)$) for every $x \in P$
  and every $y \in P'$.
  For exemple, $\nu_1$ is above $\nu_2$ but below $\nu'_1$ in permutation $\mu$.

@@

  \begin{align*}
  N_4 &= 2(2n + k + 2) + 1  = 4n + 2k + 5 \\
  N_3 &= 2(2N_4 + 2n + 2k + 4) + 1 = 20n + 12k + 9 \\
  N_2 &= 2(2N_3 + 2N_4 + 2n + 2k + 4) + 1 \\
  N_1 &= 2(2N_2 + 2N_3 + 2N_4 + 2n + 2k + 4) + 1\text{.}
  \end{align*}
  Notice that $N_1, N_2, N_3$ and $N_4$ are polynomial in $n$.
  The crucial properties are that
  (i) $N_1, N_2, N_3$ and $N_4$ are odd integers
  and
  (ii) $N_i > \left(\sum_{i < j \leq 4} 2N_j\right) + 2n + 2k + k$
  for every $1 \leq i \leq 4$.

  We now turn to defining various gadgets (sequences of integers)
  that act as building blocks in our construction:
  \begin{align*}
  \sigma'  &= ((k+1) \; \sigma \; (k+2)) \; [2N_2 + N_4 + 2n + k + 2] \\
  \pi'     &= ((n+1) \; \pi \; (n+2)) \; [2N_2 + N_4 + n + k + 2] \\
  \sigma'' &= \sigma \; [2N_2 + N_4] \\
  \pi''    &= \pi \; [2N_2 + N_4 + k] \\
  \nu_1    &= \nearrow_{N_1} \; [2N_2 + 2N_3 + 2N_4 + 2n + 2k + 4] \\
  \nu'_1   &= \nearrow_{N_1} \; [N_1 + 2N_2 + 2N_3 + 2N_4 + 2n + 2k + 4] \\
  \nu_2    &= \nearrow_{N_2} \; [N_2] \\
  \nu'_2   &= \searrow_{N_2} \\
  \nu_3    &= \nearrow_{N_3} \; [2N_2 + 2N_4 + 2n + 2k + 4] \\
  \nu'_3   &= \nearrow_{N_3} \; [2N_2 + N_3 + 2N_4 + 2n + 2k + 4] \\
  \nu_4    &= \searrow_{N_4} \; [2N_2 + N_4 + 2n + 2k + 4] \\
  \nu'_4   &= \searrow_{N_4} \; [2N_2]\text{.}
  \end{align*}
  We are now in position to define our target permutation $\mu$:
  $$
  \mu
  =
  \nu_1 \; \nu_2 \; \nu'_1 \; \nu_3 \; \sigma' \; \nu_4 \; \nu'_3 \; \pi' \; \nu'_4 \; \pi'' \; \sigma''
  \text{.}
  $$
  We claim that $\sigma$ occurs as a pattern in $\pi$ if and only if
  there exists a containment-free perfect matching
  $\mathcal{M}$ in $G_\mu$ such that
  for any two disctinct edges
  $(a, b)$ and $(a', b')$ in $\mathcal{M}$
  we have $\mu(a) < \mu(a')$ if and only if $\mu(b) < \mu(b')$.

>>>>

This section is devoted to proving hardness of recognizing square permutations.
In the same way as happens with words, we shall use a linear graph framework
where deciding whether a permutation is a square reduces to computing some
specific matching in the associated linear graph \cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013}.
It reads as follows.

\begin{lemma}
  \label{lemma:matching}
  Let $\pi$ be a permutation. The following statements are equivalent:
  \begin{enumerate}
    \item
    $\pi$ is a square.
    \item
    There exists a perfect containment-free mathing
    $\mathcal{M}$ of $G_\pi$ such that
    for any two disctinct edges
    $(a, b)$ and $(a', b')$ in $\mathcal{M}$
    we have $\pi(a) < \pi(a')$ if and only if $\pi(b) < \pi(b')$.
  \end{enumerate}
\end{lemma}

\begin{proof}[of Lemma~\ref{lemma:matching}]
    \qed
\end{proof}

Before proving hardness, we give an easy lemma that will prove extremely
useful for simplifying the proof of upcoming
Proposition~\ref{proposition:hardness}.

\begin{lemma}
  \label{lemma:at most one edge monotone}
  Let $\pi = \pi_1 \, \sigma_1 \, \pi_2 \, \sigma_2 \, \pi_3$
  be a permutation with $|\sigma_1| \geq 2$ and $|\sigma_2| \geq 2$,
  and $\mathcal{M}$ be a containment-free perfect matching of $G_\pi$.
  The following assertions hold:
  \begin{enumerate}
    \item
    If $\sigma_1$ is increasing, $\sigma_2$ is decreasing and
    $\sigma_1$ is above $\sigma_2$
    (see Fig.~\ref{subfig:increasing before and aboce decreasing}),
    then $\mathcal{M}$ sontains at most one $(\sigma_1, \sigma_2)$-edge.
    \item
    If $\sigma_1$ is decreasing, $\sigma_2$ is increasing and
    $\sigma_1$ is below $\sigma_2$
    (see Fig.~\ref{subfig:decresing before and below increasing}),
    then $\mathcal{M}$ sontains at most one $(\sigma_1, \sigma_2)$-edge.
  \end{enumerate}
\end{lemma}

@@

\begin{proof}
  The problem is certainly in $\NPclass$.
  We propose a reduction from the pattern involvement problem which is
  known to be \NPC~\cite{Bose:Buss:Lubiw:1998}:
  Given permutation $\pi$ and $\sigma$, decide whether $\sigma$
  occurs in $\pi$ (as an order-isomorphic pattern).

  For the sake of clarity, we will that a bunch of positions $P$
  is \emph{above} (resp. \emph{below}) above another bunch of positions $P'$
  in permutation $\mu$
  if $\mu(x) > \mu(y)$ (resp. $\mu(x) < \mu(y)$) for every $x \in P$
  and every $y \in P'$.
  For exemple, $\nu_1$ is above $\nu_2$ but below $\nu'_1$ in permutation $\mu$.

@@

  Let $\pi \in S_n$ and $\sigma \in S_k$ be two arbitrary permutations.
  Define
  \begin{align*}
  N_4 &= 2(2n + k + 2) + 1  = 4n + 2k + 5 \\
  N_3 &= 2(2N_4 + 2n + 2k + 4) + 1 = 20n + 12k + 9 \\
  N_2 &= 2(2N_3 + 2N_4 + 2n + 2k + 4) + 1 \\
  N_1 &= 2(2N_2 + 2N_3 + 2N_4 + 2n + 2k + 4) + 1\text{.}
  \end{align*}
  Notice that $N_1, N_2, N_3$ and $N_4$ are polynomial in $n$.
  The crucial properties are that
  (i) $N_1, N_2, N_3$ and $N_4$ are odd integers
  and
  (ii) $N_i > \left(\sum_{i < j \leq 4} 2N_j\right) + 2n + 2k + k$
  for every $1 \leq i \leq 4$.

  We now turn to defining various gadgets (sequences of integers)
  that act as building blocks in our construction of a new permutation $\mu$:
  \begin{align*}
  \sigma'  &= ((k+1) \; \sigma \; (k+2)) \; [2N_2 + N_4 + 2n + k + 2] \\
  \pi'     &= ((n+1) \; \pi \; (n+2)) \; [2N_2 + N_4 + n + k + 2] \\
  \sigma'' &= \sigma \; [2N_2 + N_4] \\
  \pi''    &= \pi \; [2N_2 + N_4 + k] \\
  \nu_1    &= \nearrow_{N_1} \; [2N_2 + 2N_3 + 2N_4 + 2n + 2k + 4] \\
  \nu'_1   &= \nearrow_{N_1} \; [N_1 + 2N_2 + 2N_3 + 2N_4 + 2n + 2k + 4] \\
  \nu_2    &= \nearrow_{N_2} \; [N_2] \\
  \nu'_2   &= \searrow_{N_2} \\
  \nu_3    &= \nearrow_{N_3} \; [2N_2 + 2N_4 + 2n + 2k + 4] \\
  \nu'_3   &= \nearrow_{N_3} \; [2N_2 + N_3 + 2N_4 + 2n + 2k + 4] \\
  \nu_4    &= \searrow_{N_4} \; [2N_2 + N_4 + 2n + 2k + 4] \\
  \nu'_4   &= \searrow_{N_4} \; [2N_2]\text{.}
  \end{align*}
  We are now in position to define our target permutation $\mu$:
  $$
  \mu
  =
  \nu_1 \; \nu_2 \; \nu'_1 \; \nu_3 \; \sigma' \; \nu_4 \; \nu'_3 \; \pi' \; \nu'_4 \; \pi'' \; \sigma''
  \text{.}
  $$

  It is immediate that H can be constructed in polynomial-time in $n$ and $k$.
  We claim that $\sigma$ occurs as a pattern in $\pi$ if and only if
  there exists a containment-free perfect matching
  $\mathcal{M}$ in $G_\mu$ such that
  for any two disctinct edges
  $(a, b)$ and $(a', b')$ in $\mathcal{M}$
  we have $\mu(a) < \mu(a')$ if and only if $\mu(b) < \mu(b')$.
